To Do List

### 2018-09-10 Manuscript/manuscript.Rmd
updates

Due to the size of Hi-C data and the abundance of available genomic features, few methods have been developed to study the role and interplay of different sets of these features on the folding of DNA. Furthermore, many accepted methods ignore key characteristics of the data that, when addressed, may improve predictive performances. One such pitfall is the issue of highly imbalanced classes. The commonality between most accepted methods for predicting TAD boundaries considers binning the genome and labeling the bins as either containing a TAD boundary or not. However, due to the sparcity of TADs throughout the genome, most bins will not contain a TAD boundary. This will contribute to highly imbalanced classes and will also introduce predictors with near-zero variance. It has been shown that imbalanced classes can result in poor predictive performances of classification algorithms and unless the class imbalance is mild, researchers should always use an appropriate method for dealing with this (Blagus et al). Another problem lies in identifying which positional characteristics of the genomic features will optimize prediction. It is unclear if distances or overlaps between genomic bins and regions of functional elements contain the most information. Lastly, given that some features tend to colocalize near TAD boundaries, most predictive models may be unable to identify the complete set of features most associated with boundary formation, and it is important that the proposed model is able to handle multiple correlated features. [??? Like what? It may be best to devote a paragraph clearly describing them. Then, describe the Mourad and other approaches; !!! stated the issues, now will discuss current methods].

@@

The types of models and the feature characteristics used in the aforementioned methods vary considerably. Additionally, a notable omission was the use of distances between regions of the functional elements and genomic bins. Because of the different cell lines available for analysis and the widening list of functional genomic elements, it can be quite useful to have a unified approach in predicting the formation of TAD boundaries. Therefore, in addition to proposing a novel data pre-processing pipeline that addresses irregularities in the data, we also apply an elastic-net model (and/or random forest?) using both distances and counts of overlaps between bins and features as predictors in our model. We aim to assess both the predictive performance of the model and the influence of genomic features on the borders that demarcate topological domains.[??? Clarity. Define 1) the problem, and 2) ways to solve it, which is modeling. Then, for modeling, describe problems connecting genomics+machine learning terms; !!!identified problem and aim]

@@

Publicly available topologically associating domain data was obtained from GEO with accession GSE53525 for the GM12878 and K562 cell lines.  The domain data was constructed from in situ Hi-C contact matrices at a 5kb and 10 kb resolution for the GM12878 and k562 cell lines respectively. The Arrowhead algorithm was used to identify TAD boundaries for a given contact matrix (Rao et. al). Identified TAD boundaries were represented by their genomic coordinates (hg19 human genome assembly), including chromosome, start and end coordinates. The start and end coordinates were concatenated, sorted, and unique coordinates representing unique TAD boundaries were obtained [??? Check for accuracy]. Next, the genome was binned into a series of 1kb intervals (flanked by 500 bases on either side of the boundary point) and an indicator vector Y was created based on whether there was a TAD boundary located in the 1kb interval (Y=1) or not (Y=0) [??? Should it be simplified - just consider boundaries to be flanked?].

@@

Annotation data was obtained from the Encyclopedia of DNA Elements (ENCODE) Consortium [??? We need a table. Check Hong 2017 supplemental, there may be something else that I don't have in GenomeRunner database. Also, check iGD data http://big.databio.org/igd/].  These annotations consisted of genomic coordinates and were used to build the feature space, X = {X1, . . ., Xp}, of the subsequent models. The data consisted of genomic coordinates and chromosomal information. For each genomic feature, an indicator variable (1 for yes; 0 for no) was created based on whether the coordinates of the annotation overlapped with the coordinates representing a flanked TAD boundary. Likewise, a continuous variable representing the distance to the center of the nearest flanked TAD boundary was included [??? Measured in?]. Genomic features that were located inside of a flanked TAD boundary were given a distance of 0. Figure X presents an illustration for how the feature space was constructed.

@@

We developed a pipeline that systematically evaluates the problems associated with genomic data in predictive modeling settings. At each stage in the pipeline, Elastic-Net models were performed and performance metrics such as ROC curves and subsequent AUCs were evaluated. The Elastic-Net model was used here because of its ability to be tuned and its computational speed relative to other machine learning algorithms. For each elastic-net model the mixing parameter, alpha, was tuned on a grid of five values (0.1, 0.325, 0.550, 0.775, and 1.0). A sequence of values for the regularization parameter, lambda, was automatically provided by the CARET package in R. The algorithm then chose the optimal combination of alpha and lambda, and the model was evaluated. [??? Unclear, is it cross-validation tuning or what?]

Prior to initiating the pipeline, predictors with near-zero variance were removed. To identify said predictors, the nearZeroVar function in the CARET package was used. The function considers two metrics. First, the ratio of the most frequent value over the second most frequent value was calculated. Ideally, this ratio would be close to one for well-behaved predictors. Next, the number of unique values divided by the total number of samples multiplied by 100 was calculated. Here, if the frequency ratio was greater than 0.99 and the percentage of unique values was less than 0.001, the predictor was flagged as having near-zero variance and removed.

>>>>

Due to the size of Hi-C data and the abundance of available genomic features, few methods have been developed to study the role and interplay of different sets of these features on the folding of DNA. Furthermore, many accepted methods ignore key characteristics of the data that, when addressed, may improve predictive performances. One such pitfall is the issue of highly imbalanced classes. The commonality between most accepted methods for predicting TAD boundaries considers binning the genome and labeling the bins as either containing a TAD boundary or not. However, due to the sparcity of TADs throughout the genome, most bins will not contain a TAD boundary. This will contribute to highly imbalanced classes and will also introduce predictors with near-zero variance. It has been shown that imbalanced classes can result in poor predictive performances of classification algorithms and unless the class imbalance is mild, researchers should always use an appropriate method for dealing with this (Blagus et al). Another problem lies in identifying which positional characteristics of the genomic features will optimize prediction. It is unclear if distances or overlaps between genomic bins and regions of functional elements contain the most information. Lastly, given that some features tend to colocalize near TAD boundaries, most predictive models may be unable to identify the complete set of features most associated with boundary formation, and it is important that the proposed model is able to handle multiple correlated features.

@@

The types of models and the feature characteristics used in the aforementioned methods vary considerably. Additionally, a notable omission was the use of distances between regions of the functional elements and genomic bins. Because of the different cell lines available for analysis and the widening list of functional genomic elements, it can be quite useful to have a unified approach in predicting the formation of TAD boundaries. Therefore, in addition to proposing a novel data pre-processing pipeline that addresses irregularities in the data, we also apply an elastic-net model (and/or random forest?) using both distances and counts of overlaps between bins and features as predictors in our model. We aim to assess both the predictive performance of the model and the influence of genomic features on the borders that demarcate topological domains.[??? Clarity. Define 1) the problem, and 2) ways to solve it, which is modeling. Then, for modeling, describe problems connecting genomics+machine learning terms]

@@

Publicly available topologically associating domain data was obtained from GEO with accession GSE53525 for the GM12878 and K562 cell lines.  The domain data was constructed from in situ Hi-C contact matrices at a 5kb and 10 kb resolution for the GM12878 and k562 cell lines respectively. The Arrowhead algorithm was used to identify TAD boundaries for a given contact matrix (Rao et. al). Identified TAD boundaries were represented by their genomic coordinates (hg19 human genome assembly), including chromosome, start and end coordinates. The start and end coordinates were concatenated, sorted, and unique coordinates representing unique TAD boundaries were obtained [??? Check for accuracy; !!! what do you mean by accuracy here?]. The establishment of the data for modeling first consisted of binning the genome into equal sized intervals. This was performed at the same resolution as the respective cell line. That is, for GM12878, 5 kb genomic bins were used, while 10 kb intervals were used for K562. Next, for the domain data provided by each cell line, the boundary coordinates were flanked on either side by 5 kb [!!! use the same size flank for k562? (resolution was 10kb)]. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing an boundary (Y=0).[??? Should it be simplified - just consider boundaries to be flanked?; !!!].

@@

Annotation data was obtained from the Encyclopedia of DNA Elements (ENCODE) Consortium [??? We need a table. Check Hong 2017 supplemental, there may be something else that I don't have in GenomeRunner database. Also, check iGD data http://big.databio.org/igd/; !!! table in supplementary file, Hong considers TFBS, should we also include?].  These annotations consisted of genomic coordinates and were used to build the feature space, X = {$X_{1}$, . . ., $X_{p}$}, of the subsequent models. The full list of features is provided in [X]. For each genomic feature, two predictors types were considered, a count variable and a distance variable. The count variable was based on the number of regions of each functional element that overlapped with a particular genomic bin. The distance variable consisted of the length (in base pairs) from the center of each genomic bin to the center of the nearest region of a functional element. If the center of the region given by a particular functional element was located within a genomic bin, then the distance was given to be 0.[??? Measured in?; !!!]. Figure X presents an illustration for how both the response vector and feature space were constructed.

@@

We developed a pipeline that systematically evaluates the problems associated with genomic data in predictive modeling settings.

At each stage in the pipeline, Elastic-Net models were performed and performance metrics such as ROC curves and subsequent AUCs were evaluated. The Elastic-Net model was used here because of its ability to be tuned and its computational speed relative to other machine learning algorithms. For each elastic-net model the mixing parameter, alpha, was tuned on a grid of five values (0.1, 0.325, 0.550, 0.775, and 1.0). A sequence of values for the regularization parameter, lambda, was automatically provided by the CARET package in R. The algorithm then chose the optimal combination of alpha and lambda, and the model was evaluated. [??? Unclear, is it cross-validation tuning or what?]

Prior to initiating the pipeline, predictors with near-zero variance were removed. To identify said predictors, the nearZeroVar function in the CARET package was used. The function considers two metrics. First, the ratio of the most frequent value over the second most frequent value was calculated. Ideally, this ratio would be close to one for well-behaved predictors. Next, the number of unique values divided by the total number of samples multiplied by 100 was calculated. Here, if the frequency ratio was greater than 20 and the percentage of unique values was less than 10, the predictor was flagged as having near-zero variance and removed.

### 2018-09-10 ToDo.Rmd
updates

## 09/04/18

* use width in base pairs as type of feature instead of counts of overlap
   - distance and width in base pairs of overlap as features
* finish methods

>>>>

## 09/07/18

####To-Do
+ use width in base pairs as type of feature instead of counts of overlap
   - distance and width in base pairs of overlap as features
+ finish introduction, start on methods

## 09/10/18

####Questions

* Should we use elastic-net and RF as final models or just one or the other?
* In results section should we compare between mourad models visually or verbally or none at all?
* I've mentioned other methods in intro, it will be hard to compare to the others; how to handle this?
* Hong et al uses TFBS should I also include them?

####To-Do

*Finish Methods
*Finish running code applied to 100 iterations of undersampling
*Apply pipeline to K562 cell line
